Complexity analysis and algorithms for the college application problem

### 2022-06-28 paper/paper.tex
init short paper

\newif\ifen\enfalse
% Language selector; must be on line 3 for compile.sh to work
% Change to false to export Korean

\documentclass[11pt]{article} %try amsproc, amsart
\usepackage{kotex}
\usepackage{geometry}       \geometry{a4paper, margin=2.5cm}  
\usepackage{setspace}       \setstretch{1.1}
\usepackage{hyperref}       \hyperbaseurl{}    
                            \urlstyle{same}
\usepackage[ruled,linesnumbered]{algorithm2e}   \SetKwInOut{KwParams}{Parameters}
                                                \newcommand{\lIfElse}[3]{\lIf{#1}{#2 \textbf{else}~#3}}
\usepackage{stackengine}    \newcommand\xrowht[2][0]{\addstackgap[.5\dimexpr#2\relax]{\vphantom{#1}}}
\usepackage{fancyhdr}
\usepackage{graphicx}

% Author block
\usepackage{authblk}        \providecommand{\keywords}[1]{\textbf{Keywords:\ } #1}
\ifen
    \newcommand{\thetitle}{The College Application Problem}
    \newcommand{\therunningtitle}{\thetitle}
    \newcommand{\thedate}{\today}
    \newcommand{\theauthor}{Max Kapur}
    \newcommand{\theotherauthor}{Sung-Pil Hong}
\else
    \newcommand{\thetitle}{대학 입학 지원 최적화 문제}
    \newcommand{\therunningtitle}{\thetitle}
    \newcommand{\thedate}{\today}
    \newcommand{\theauthor}{Max Kapur}
    \newcommand{\theotherauthor}{홍성필}
\fi

\title{\thetitle}
\author{\theauthor}
\date{\thedate}

\ifen
\affil{Seoul National University\\
Department of Industrial Engineering\\
Management Science/Optimization Lab\\
Correspondence: \url{maxkapur@gmail.com}\\
Advisor: \theotherauthor}
\else
\affil{서울대학교\\
산업공학과 \\
경영과학/최적화 연구실\\
이메일: \url{maxkapur@gmail.com}\\
지도 교수: \theotherauthor}
\fi

% Header and footer
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0pt}
\fancyhf{}
\lhead{\theauthor}
\chead{\thedate}
\rhead{\thetitle}
\cfoot{\thepage}

@@

\ifen
    \newtheorem{theorem}{Theorem}
    \newtheorem{lemma}{Lemma}
    \newtheorem{corollary}{Corollary}
    \newtheorem{proposition}{Proposition}
    \theoremstyle{definition}
    \newtheorem{example}{Example}
    \newtheorem{definition}{Definition}
    \newtheorem{problem}{Problem}
    \newtheorem{assumption}{Assumption}
    \newtheorem{conjecture}{Conjecture}
\else
    \newtheorem{theorem}{정리}
    \newtheorem{lemma}{기본 정리}
    \newtheorem{corollary}{따름정리}
    \newtheorem{proposition}{제의}
    \theoremstyle{definition}
    \newtheorem{example}{예}
    \newtheorem{definition}{정의}
    \newtheorem{problem}{문제}
    \newtheorem{assumption}{가정}
    \newtheorem{conjecture}{추측}
\fi

\ifen\else
    \renewcommand*{\proofname}{증명\@} 
    \renewcommand{\figurename}{그림}
    \renewcommand{\tablename}{표}
\fi

@@

% Unused (I think)
%\usepackage{times}
%\usepackage{letltxmacro}
%\usepackage[shortlabels]{enumitem}

@@

\begin{abstract}
This paper considers the maximization of the expected maximum value of a portfolio of random variables subject to a budget constraint. We refer to this as the optimal college application problem. When each variable's cost, or each college's application fee, is identical, we show that the optimal portfolios are nested in the budget constraint, yielding an exact polynomial-time algorithm. When colleges differ in their application fees, we show that the problem is NP-complete. We provide four algorithms for this more general setup: a branch-and-bound routine, a dynamic program that produces an exact solution in pseudopolynomial time, a different dynamic program that yields a fully polynomial-time approximation scheme, and a simulated-annealing heuristic. Numerical experiments demonstrate the algorithms' accuracy and efficiency.

\begin{keywords}
submodular maximization, knapsack problems, approximation algorithms
\end{keywords}

\ifen\else The full text of this paper is available in English at \url{https://github.com/maxkapur/CollegeApplication}.\fi
\end{abstract}

\renewcommand{\abstractname}{요약}
\begin{abstract}
본 논문은 다수의 확률 변수로 구성된 포트폴리오의 기대 최댓값을 예산 조건 하에서 최대화하는 문제를 고려한다. 이를 대학 입학 지원 최적화 문제라고 부른다. 각 확률 변수의 비용, 즉 각 대학의 지원 비용이 동일한 경우, 최적 포트폴리오는 예산 제약식으로 결정된 포함 사슬 관계 성질을 가짐을 보이고 이를 바탕으로 다항 시간 해법을 제시한다. 대학의 지원 비용이 서로 다른 경우, 문제가 NP-complete함을 증명한다. 일반적인 문제를 위해 4가지 해법을 제시하며, 분지한계 기반 해법, 의사 다항 시간에 정확한 해를 출력하는 동적 계획 해법, 다른 동적 계획 기반으로 완전 다항 시간 근사 해법(fully polynomial-time approximation scheme), 그리고 모의 담금질(simulated annealing)을 이용한 휴리스틱 해법을 포함한다.
% 첫째는 분지한계 기반 해법이다. 둘째는 의사 다항 시간에 정확한 해를 출력하는 동적 계획 해법이다. 마지막으로 다른 동적 계획 기반으로 완전 다항 시간 근사 해법(fully polynomial-time approximation scheme)을 도출한다.
수리적 실험을 통해 알고리즘의 정확도와 효율성을 보인다.

\ifen 본 논문은 \url{https://github.com/maxkapur/CollegeApplication}에서 전문을 한글로 공유한다. \fi
\end{abstract}

@@

\pagebreak
\ifen \section{Introduction}  \else \section{서론} \fi
\ifen This paper considers the following optimization problem:
\else 본 논문은 다음과 같은 최적화 문제를 고려한다.\fi
%\begin{definition}[Optimal college application portfolio] \label{generalproblemstatement}
\begin{align} \label{headlineproblem}
\begin{split}
\text{maximize}\quad & v(\mathcal{X}) =  \operatorname{E}\Bigl[\max\bigr\{t_0,
\max\{t_j Z_j : j \in \mathcal{X}\}\bigr\}\Bigr] \\
\text{subject to}\quad & \mathcal{X} \subseteq \mathcal{C}, ~~\sum_{j\in \mathcal{X}} g_j \leq H
\end{split}
\end{align}
%\end{definition}
\ifen Here $\mathcal{C} = \{ 1 \dots m\}$ is an index set; $H > 0$ is a budget parameter; for $j = 1 \dots m$, $g_j > 0$ is a cost parameter and $Z_j$ is a random, independent Bernoulli variable with probability $f_j$; and for $j = 0\dots m$, $t_j\geq 0$ is a utility parameter. 
\else 단,  $\mathcal{C} = \{ 1 \dots m\}$은 지표 집합이며 $H$는 예산을 나타내는 모수다. 각 $j = 1 \dots m$에 대해 $g_j > 0$는 비용 모수이며 $Z_j$는 확률 $f_j$를 가지는 서로 독립적인 Bernoulli 변수다. 각 $j = 0 \dots m$에 대해 $t_j\geq 0$는 효용 모수다. \fi

\ifen
We refer to this problem as the \emph{optimal college application} problem, as follows. Consider an admissions market with $m$ colleges. The $j$th college is named $c_j$. Consider a single prospective student in this market, and let each $t_j$-value indicate the utility she associates with attending $c_j$, where her utility is $t_0$ if she does not attend college. Let $g_j$ denote the application fee for $c_j$ and $H$ the student's total budget to spend on application fees. Lastly, let $f_j$ denote the student's probability of being admitted to $c_j$ if she applies, so that $Z_j$ equals one if she is admitted and zero if not. It is appropriate to assume that the $Z_j$ are statistically independent as long as $f_j$ are probabilities estimated specifically for this student (as opposed to generic acceptance rates). Then the student's objective is to maximize the expected utility associated with the best school she gets into within this budget. Therefore, her optimal college application strategy is given by the solution $\mathcal{X}$ to the problem above, where $\mathcal{X}$ represents the set of schools to which she applies. 
\else
다음 해석에 따라 이를 `대학 입학 지원 최적화 문제'라고 부른다. $m$개의 대학교를 가지는 입학 시장을 고려하자. $j$번째 학교의 이름은 $c_j$다. 어떤 학생이 $c_j$에 입학하게 되면 효용 $t_j$를 얻게 된다고 하자. 단, 어떤 대학에도 진학하지 않는 경우 그 학생의 효용은 $t_0$이다. $c_j$의 지원 비용이 $g_j$이며 학생이 지원에 쓸 수 있는 예산이 $H$라고 하자. 마지막으로, $c_j$에 지원하면 학생이 합격할 확률이 $f_j$이라고 하자. 따라서 합격하면 $Z_j = 1$, 합격 안 하면 $Z_j = 0$이 된다. $f_j$가 (학교의 전체적인 합격률이 아니라) 바로 이 학생의 합격 확률일 때 $Z_j$의 확률적 독립성은 적절한 가정이다. 그러면 학생의 목표는 주어진 예산 안에서 학생이 합격하는 학교들에서 얻는 효용 중 기대 최댓값을 최대화하는 것이다. 위 문제의 최적해가 $\mathcal{X}$일 때, 학생의 최적 대학 지원 전략은 $\mathcal{X}$에  속한 학교로 지원하는 것이다.
\fi

\ifen
The college application problem is not solely of theoretical interest. Due to the widespread perception that career earnings depend on college admissions outcomes, the solution of \eqref{headlineproblem} holds monetary value. In the American college consulting industry, students pay private consultants an average of \$200 per hour for assistance in preparing application materials, estimating their admissions odds, and identifying target schools (Sklarow 2018). In Korea, an important revenue stream for admissions consulting firms such as Megastudy (\url{megastudy.net}) and Jinhak (\url{jinhak.com}) is ``mock application'' software that claims to use artificial intelligence to optimize the client's application strategy. However, although predicting admissions outcomes on a school-by-school basis is a standard benchmark for logistic regression models (Acharya et al. 2019), we believe our study is the first to focus on the overall application strategy as an optimization problem.

The problem is also conformable to other competitive matching games such as job application. Here, the budget constraint may represent the time needed to complete each application, or a legal limit on the number of applications permitted.
\else
대학 입학 지원 최적화 문제는 이론적인 화제일 뿐만이 아니다. 미래의 소득이 대학 입학 결과로 결정된다는 넓은 인식에 따라, \eqref{headlineproblem}의 최적해는 금전적인 가치를 가지고 있다. 미국 입학 컨설팅 산업에서 대학 지원 서류 작성, 합격 확률 추정, 그리고 지원 학교를 선택하는 데에 자문하는 개인 상담가의 시간당 급료는 평균 200달러다 (Sklarow 2018). 한국에서, 메가스터디(\url{megastudy.net})와 진학(\url{jinhak.com})과 같은 입학 컨설팅 기업의 주된 수입원 중 인공지능을 활용하여 학생의 지원 전략을 최대화한다고 홍보하는 ``모의 지원'' 소프트웨어가 있다. 그러나, 학교 개별 합격 확률을 추정하는 것은 로지스틱 회귀 분석 모형의 익숙한 응용 사례(Acharya et al. 2019)이지만, 전체적인 지원 전략에 집중하여 최적화 문제로 모형화한 것은 본 연구의 새로운 기여로 보인다.

취직 전략과 같은 유사한 경쟁적 매칭 게임에도 위의 문제를 적용할 수 있다. 이때, 예산 제약 조건은 원서를 작성하는 시간이나 지원 개수에 대한 법적인 제한을 나타낼 수 있다.
\fi

\ifen \subsection{Methodological orientation} \else \subsection{방법록적 지향} \fi
\ifen
The college application problem straddles several methodological universes. Its stochastic nature recalls classical portfolio allocation models. However, the knapsack constraint renders the problem NP-complete, and necessitates combinatorial solution techniques. We also observe that the objective function is also a submodular set function, although our approximation results suggest that college application is a relatively easy instance of submodular maximization.
\else
대학 지원 문제는 다양한 방법론적인 우주를 걸치는 문제다.  확률적인 문제인 만큼 재정학 분야에 뿌리를 가지는 포트폴리오 배분 모형과 비슷하다. 그러나 배낭 제약 조건은 대학 지원 문제를 NP-complete하게 만들며 조합적인 해법이 필요하다. 목적함수는 또한 submodular 집합 함수이지만, 본 연구가 제시하는 근사 해법 결과는 대학 지원 문제가 일반 submodular 함수 최대화 문제의 비교적 쉬운 경우임으로 해석할 수 있다.
\fi

% reword this sentence for a better transition
\ifen
% In an equilibrium analysis of the American college market, Fu (2014) remarked that
% (and change below)
A problem similar to \eqref{headlineproblem} arose in an equilibrium analysis of the American college market by Fu (2014), who described college application as a ``nontrivial portfolio problem'' (226). In computational finance, traditional portfolio allocation models weigh the sum of expected profit across all assets against a risk term, yielding a concave maximization problem with linear constraints (Markowitz 1952; Meucci 2005). But college applicants maximize the expected value of their \emph{best} asset: If a student is admitted to her $j$th choice, then she is indifferent as to whether she gets into her $(j+1)$th choice. As a result, student utility is \emph{convex} in the utility associated with individual applications. Risk management is implicit in the college application problem because, in a typical admissions market, college preferability correlates inversely with competitiveness. That is, students negotiate a tradeoff between attractive, selective “reach schools” and less preferable “safety schools” where admission is a safer bet (Jeon 2015). Finally, the combinatorial nature of the college application problem makes it difficult to solve using the gradient-based techniques associated with continuous portfolio optimization. Fu estimated her equilibrium model (which considers application as a \emph{cost} rather than a constraint) by clustering the schools so that $m=8$, a scale at which enumeration is tractable. We pursue a more general solution.
\else
미국 대학 입학 시장의 균형 분석에서, Fu (2014)는 \eqref{headlineproblem}과(와) 비슷한 부문제를 직면하였으며 이를 ``nontrivial portfolio problem''이라고 부른다 (226). 재정학 분야에서, 고전적 포트폴리오 배분 최적화 모형은 전체 자산에 대한 기대 총이익에서 위험회피 항을 뺌으로 선형식으로 제약된 오목 최대화 문제를 이룬다 (Markowitz 1952; Meucci 2005). 그러나 대학 지원자는 가치가 제일 높은 단일 자산의 기대 가치를 최대화하고자 한다. 어떤 학생이 자신이 $j$번째로 선호하는 학교에 합격하면 $(j+1)$번째 학교의 합격 여부는 무관한 상황이 된다. 이는 학생의 효용을 각 지원 발송의 효용에 대해 오목이 아닌 볼록 함수로 만든다. 또한 입학 시장에서는 전형적으로 대학의 효용과 합격 확률이 서로 반비례하므로 대학 지원 문제는 위험 관리를 포함하게 된다. 특히 선호도가 높으며 붙기 어려운 “상향” 지원 학교(reach school)와 선호도가 낮으며 붙기 쉬운 “안정” 지원 학교(safety school) 사이의 균형을 고려해야 한다 (전민희 2015). 마지막, 대학 지원의 조합적인 본질로 인해 연속적인 포트폴리오 최적화 문제에서 흔히 사용하는 기울기 해법으로는 풀기가 어렵다. Fu는 지원 비용을 제약 조건 대신 목적함수의 한 항으로 모형화했으며, 균형 모형의 모수를 추정하기 위해 $m=8$이 되도록 학교들의 클러스터를 먼저 구성했다. 이는 열거법을 통해 문 쉽게 풀 수 있는 규모이지만 본 연구는 더 일반적인 $m$에 대한 해법을 추구한다.
\fi

\ifen
The integer formulation of the college application problem can be viewed as a kind of binary knapsack problem with a polynomial objective function of degree $m$. Our branch-and-bound and dynamic programming algorithms closely resemble existing algorithms for knapsack problems (Martello and Toth 1990, \S\,2.5--6). In fact, by manipulating the admissions probabilities, the objective function can be made to approximate a linear function of the characteristic vector to an arbitrary degree of accuracy, a fact that we exploit in our NP-completeness proof. Previous research has introduced various forms of stochasticity to the knapsack problem, including variants in which each item's utility takes a known probability distribution (Steinberg and Parks 1979; Carraway et al. 1993) and an online context in which the weight of each item is observed after insertion into the knapsack (Dean et al. 2008). Our problem superficially resembles the preference-order knapsack problem considered by Steinberg and Parks and Carraway et al., but these models lack the college application problem's singular ``maximax'' form. Additionally, unlike those models, we do not attempt to replace the real-valued objective function with a preference order over \emph{outcome distributions,} which introduces technical issues concerning competing notions of stochastic dominance (Sniedovich 1980). We take for granted the student's preferences over \emph{outcomes} (as encoded in the $t_j$-values), and focus instead on an efficient computational approach to the well-defined problem above.
\else
대학 지원 문제의 정수 모형은 $m$차 다항식을 목적함수로 갖춘 일종의 이진 배낭 문제로 볼 수 있다. 본 논문에서 제시하는 분지한계법과 동적 계획 해법은 배낭 문제를 위한 기존 알고리즘과 매우 비슷하다 (Martello와 Toth 1990, \S\,2.5--6). 입학 확률을 적당히 조정하면 특성벡터의 선형 함수를 원래 목적함수로 원하는 만큼의 정확성을 가지고 근사할 수 있으며 NP-completeness 증명에서 이 성질을 활용한다. 배낭 문제에 확률성을 도입한 선행 연구 중, 각 상품의 효용이 정해진 확률 분포로 결정되는 모형(Steinberg와 Parks 1979; Carraway 외 1993), 그리고 배낭에 삽입한 다음에 상품의 무게를 관측할 수 있는 온라인 모형 (Dean 외 2008) 등이 있다. 본 연구가 고려하는 문제는 Steinberg와 Parks 그리고 Carraway 외가 고려한 ``우선순위 배낭 문제''와는 유사성을 가지지만, 우선순위 배낭 문제는 대학 지원 문제의 특색인 `maximax' 형태를 가지지는 않는다. 또한 우선순위 모형과 달리, 본 연구에서 실수값 목적함수를 선호 순위를 정의해야 하는 `결과 분포'로 대체할 필요가 없다. 확률적 우위에 대한 상반된 개념들의 문제를 야기시키기 때문이다 (Sniedovich 1980). 대신 $t_j$-값으로 유도된 학생의 `결과'에 대한 선호 순위를 받아들여 위에서 정의한 것처럼 명확히 정의된 문제를 위한 효율적인 계산법을 지향한다.
\fi

>>>>

\documentclass[11pt]{article} %try amsproc, amsart
\usepackage{geometry}       \geometry{a4paper, margin=2.5cm}  
\usepackage{hyperref}       \hyperbaseurl{}    
                            \urlstyle{same}
\usepackage[ruled,linesnumbered]{algorithm2e}   \SetKwInOut{KwParams}{Parameters}
                                                \newcommand{\lIfElse}[3]{\lIf{#1}{#2 \textbf{else}~#3}}
\usepackage{stackengine}    \newcommand\xrowht[2][0]{\addstackgap[.5\dimexpr#2\relax]{\vphantom{#1}}}

% Author block
\usepackage{authblk}

\title{Complexity analysis and algorithms for the college application problem}
\author[1]{Max Kapur}
\author[2]{Sung-Pil Hong}
\affil[1,2]{Department of Industrial Engineering, Seoul National University}
\affil[1]{Email: maxkapur@gmail.com}
\date{\today}

@@

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}
\newtheorem{problem}{Problem}
\newtheorem{assumption}{Assumption}
\newtheorem{conjecture}{Conjecture}

@@

@@

@@

\section{Introduction}

This study considers the following optimization problem:
\begin{align} \label{headlineproblem}
\begin{split}
\text{maximize}\quad & % v(\mathcal{X}) =
 \operatorname{E}\Bigl[\max\bigr\{t_0,
\max\{t_j Z_j : j \in \mathcal{X}\}\bigr\}\Bigr] \\
\text{subject to}\quad & \mathcal{X} \subseteq \mathcal{C}, ~~\sum_{j\in \mathcal{X}} g_j \leq H
\end{split}
\end{align}
Here $\mathcal{C} = \{ 1 \dots m\}$ is an index set; $H > 0$ is a budget parameter; for $j = 1 \dots m$, $g_j > 0$ is a cost parameter and $Z_j$ is a random, independent Bernoulli variable with probability $f_j$; and for $j = 0\dots m$, $t_j\geq 0$ is a utility parameter.

We refer to this problem as the \emph{optimal college application} problem, as follows. Consider an admissions market with $m$ colleges. The $j$th college is named $c_j$. Consider a single prospective student in this market, and let each $t_j$-value indicate the utility she associates with attending $c_j$, where her utility is $t_0$ if she does not attend college. Let $g_j$ denote the application fee for $c_j$ and $H$ the student's total budget to spend on application fees. Lastly, let $f_j$ denote the student's probability of being admitted to $c_j$ if she applies, so that $Z_j$ equals one if she is admitted and zero if not. It is appropriate to assume that the $Z_j$ are statistically independent as long as $f_j$ are probabilities estimated specifically for this student (as opposed to generic acceptance rates). Then the student's objective is to maximize the expected utility associated with the best school she gets into within this budget. Therefore, her optimal college application strategy is given by the solution $\mathcal{X}$ to the problem above, where $\mathcal{X}$ represents the set of schools to which she applies.

% This problem has applications in other competitive matching games. However, we believe we are the rist

\section{Preliminaries}

For the remainder of the paper, we assume with minimal loss of generality that $f_j \in \mathbb{Q}$, $t_j \in \mathbb{N}$, $g_j \in \mathbb{N}$, and $H \in \mathbb{N}$; that the $f_j$-values have the same denominator $D$; and that $t_0 < t_1 \leq \dots \leq t_m$, $g_j \leq H$, and $\sum g_j > H$. Unless otherwise noted, we assume that $t_0 = 0$, an assumption we justify below.

\subsection{The objective function}

This subsection derives a closed-form expression for the objective function of \eqref{headlineproblem}.

We refer to the set $\mathcal{X} \subseteq \mathcal{C}$ of schools to which a student applies as her \emph{application portfolio.} The expected utility the student receives from $\mathcal{X}$ is called its \emph{valuation}. Given an application portfolio, let $p_j(\mathcal{X})$ denote the probability that the student attends $c_j$. This occurs if and only if she \emph{applies} to $c_j$, is \emph{admitted} to $c_j$, and is \emph{rejected} from any school she prefers to $c_j$; that is, any school with higher index. Hence, for $j= 0\dots m$,
\begin{align}
p_j(\mathcal{X}) &= 
\begin{cases}
\displaystyle f_j  \prod_{\substack{i \in \mathcal{X}: \\ i > j}} (1 - f_{i}), \quad & j \in \{0\}\cup\mathcal{X}\\
0, \quad & \text{otherwise}
\end{cases} 
\end{align}

The following proposition follows by computing $v(\mathcal{X}) = \sum_{j=0}^m  t_j p_j(\mathcal{X})$.

\begin{proposition}[Closed form of portfolio valuation function]
\begin{align}
v(\mathcal{X}) &= \sum_{j=0}^m t_j p_j(\mathcal{X}) = \sum_{j\in\{0\}\cup\mathcal{X}} \Bigl( f_j t_j \prod_{\substack{i \in \mathcal{X}: \\ i > j}} (1 - f_{i}) \Bigr)  \label{closedformportfoliovaluationX}%, \quad \text{or equivalently,}\\
%\qquad v(x) &= t_0 \prod_{j=1}^m (1 - f_{j} x_j) + \sum_{j=1}^m \Bigl( x_j t_j f_j \prod_{j’ = j+1}^m (1 - f_{j’} x_{j’}) \Bigr) \label{closedformportfoliovaluationx}
\end{align}
\end{proposition}
%\begin{proof}Computing $v(\mathcal{X}) = \sum_{j=0}^m  t_j p_j(\mathcal{X})$ yields \eqref{closedformportfoliovaluationX}. Next, because $1 - f_j x_j = 1$ if $x_j = 0$, we may define $p_j$ equivalently as $p_j(x) = x_j  f_j \prod_{j’ = j+1}^m (1 - f_{j’} x_{j’})$ to obtain \eqref{closedformportfoliovaluationx}. 
%\end{proof}

Next, we show that without loss of generality, we may assume that $t_0 = 0$.

\begin{lemma} \label{assumetzerozero}
For some $\gamma \leq t_0$, let $\bar t_j = t_j - \gamma$ for $j = 0 \dots m$. Then $v(\mathcal{X}; \bar t_j) = v(\mathcal{X};  t_j) -  \gamma$ regardless of $\mathcal{X}$. 
\end{lemma}
\begin{proof}
By definition, $\sum_{j=0}^m p_j(\mathcal{X}) = \sum_{j \in \{0\}\cup\mathcal{X}} p_j(\mathcal{X}) = 1$. Therefore
\begin{align}
\begin{split}
v(\mathcal{X}; \bar t_j) &= \sum_{j\in \{0\}\cup\mathcal{X}}  \bar t_j p_j(\mathcal{X})
=\sum_{j\in \{0\}\cup\mathcal{X}} (t_j - \gamma) p_j(\mathcal{X}) \\
&=\sum_{j\in \{0\}\cup\mathcal{X}} t_j p_j(\mathcal{X})  - \gamma 
= v(\mathcal{X}; t_j) - \gamma
\end{split} 
\end{align}
which completes the proof.
\end{proof}

\subsection{Submodularity}

%Now, we show that the portfolio valuation function is submodular. This result is primarily of taxonomical interest and may be safely skipped. Our solution algorithms for the college application problem are better than generic algorithms for submodular maximization, and we prove their validity using elementary analysis.

\begin{definition}[Submodular set function]
Given a ground set $\mathcal{C}$ and function $v : 2^{\mathcal{C}} \mapsto \mathbb{R}$, $v(\mathcal{X})$ is called a \emph{submodular set function} if and only if $v(\mathcal{X}) + v(\mathcal{Y}) \geq v(\mathcal{X}\cup\mathcal{Y}) + v(\mathcal{X}\cap\mathcal{Y})$
for all $\mathcal{X}, \mathcal{Y} \subseteq \mathcal{C}$. Furthermore, if $ v(\mathcal{X}\cup\{k\}) - v(\mathcal{X}) \geq 0$ for all $\mathcal{X} \subset \mathcal{C}$ and $k \in \mathcal{C} \setminus \mathcal{X}$, $v(\mathcal{X})$ is said to be a \emph{nondecreasing} submodular set function.
\end{definition}

\begin{theorem}
The college application portfolio valuation function
$v(\mathcal{X})$ % = \sum_{j\in\mathcal{X}} \Bigl( f_j t_j \prod_{\substack{i \in \mathcal{X}: \\ i > j}} (1 - f_{i}) \Bigr)\]
is a nondecreasing submodular set function.
\end{theorem}

\begin{proof}
See Kapur (2022, \S\,2.3). 
\end{proof}

\section{NP-completeness}

In this section, we will show that the college application problem is NP-complete by transformation from the binary knapsack problem, which is known to be NP-complete (Garey and Johnson 1979, \S\,3.2.1). We begin by presenting the knapsack and college applications as decision problems.

\begin{definition}[Decision form of knapsack problem (KP)]
\textbf{Instance:} a set $\mathcal{B}$ of $m$ objects, utility values $u_j \in \mathbb{N}$ and weight $w_j \in \mathbb{N}$ for each $j \in \mathcal{B}$, knapsack capacity $W\in \mathbb{N}$, and target utility $U\in \mathbb{N}$. \textbf{Question:} Is there a set $\mathcal{B’} \subseteq \mathcal{B}$ having $\sum_{j \in \mathcal{B’}} u_j \geq U$ and  $\sum_{j \in \mathcal{B’}} w_j \leq W$?
\end{definition}

\begin{theorem}
KP is NP-complete. 
\end{theorem}

\begin{proof}See Garey and Johnson (1979, \S\,3.2.1).\end{proof}

\begin{definition}[Decision form of the college application problem (CAP)] \label{ellisdecisionform}
\textbf{Instance:} an instance $(f, t, g, H)$ of the college application problem and a target valuation $V$. \textbf{Question:} Is there a a portfolio $\mathcal{X} \subseteq \mathcal{C}$ having $v(\mathcal{X}) \geq V$ and  $\sum_{j \in \mathcal{X}} g_j \leq H$?
\end{definition}

\begin{theorem}
CAP is NP-complete.
\end{theorem}

\begin{proof}
(CAP $\in$ NP.) Self-evident. % Let $D$ denote
%Observe that because $v(\mathcal{X})$ is the product of $m$ rational numbers having denominator $D$, $v(\mathcal{X})$ itself can be expressed as a rational number with denominator $D^m$. 
%let \[D= \max\bigl\{\max t_j, \max g_j, \operatorname{LCD}(f_j)\bigr\}\] (recall that $t_j, g_j \in \mathbb{N}$). Then problem data can be written in $O(D m)$-space, and $v(\mathcal{X}) \in \mathbb{Q}$. Its numerator is at most $(D^m)^2$ and its denominator is at most $D^m$. Therefore, a verifier for $\mathcal{X}$ needs \[O\bigl((\log D^m)^2 \bigr) = O\bigl((m \log D)^2 \bigr) \subset O(D^2 m^2 )\] bits of information in order to calculate $v(\mathcal{X})$, which is polynomial in the problem size. If $f_j \in \mathbb{R}$, then the problem is NP-hard rather than NP-complete.

(KP $\propto$ CAP.) Consider an instance of the knapsack problem, and we will construct an instance of Problem \ref{ellisdecisionform} that is a yes-instance if and only if the corresponding knapsack instance is a yes-instance. Without loss of generality, we may assume that the objects in $\mathcal{B}$ are indexed in increasing order of $u_j$, that each $u_j >0$, and that each $w_j \leq W$.

Let $U_{\mathrm{max}} = \sum_{j \in \mathcal{B}} u_j$ and $\delta = {1} /{m U_{\mathrm{max}}} > 0$, and construct an instance of CAP with $\mathcal{C} = \mathcal{B}$, $H = W$, all $f_j = \delta$, and each $t_j = u_j / \delta$. Clearly, $\mathcal{X} \subseteq \mathcal{C}$ is feasible for CAP if and only if it is feasible for the knapsack instance. Now, we observe that for any nonempty $\mathcal{X}$,

\begin{align}
\begin{split}
\sum_{j \in \mathcal{X}} u_j &=  \sum_{j \in \mathcal{X}} f_j t_j
> \sum_{j \in \mathcal{X}} \Bigl( f_j t_j \prod_{\substack{j’ \in \mathcal{X}: \\ j' > j}} (1 - f_{j’}) \Bigr)
= v(\mathcal{X}) \\
&= \sum_{j \in \mathcal{X}} \Bigl( u_j  \prod_{\substack{j’ \in \mathcal{X}: \\ j' > j}} (1 - \delta) \Bigr)
\geq (1 - \delta)^m \sum_{j \in \mathcal{X}} u_j \\
&\geq (1 - m\delta) \sum_{j \in \mathcal{X}} u_j 
\geq \sum_{j \in \mathcal{X}} u_j  - m\delta U_{\mathrm{max}}
= \sum_{j \in \mathcal{X}} u_j  - 1.
\end{split}
\end{align}

This means that the utility of an application portfolio $\mathcal{X}$ in the corresponding knapsack instance is the smallest integer greater than $v( \mathcal{X})$. That is, $\sum_{j \in \mathcal{X}} u_j \geq U$ if and only if $v(\mathcal{X}) \geq U -1$. Taking $V = U-1$ completes the transformation and concludes the proof. 
\end{proof}

\section{Optimization algorithms}

In this section, we present three optimization algorithms for the college application problem. The first two are exact dynamic programming (DP) algorithms. The third algorithm is a fully polynomial-time approximation scheme (FPTAS) based on the second DP.

 % We will use this algorithm as the basis for an FPTAS for the college application problem that uses $O(m^3 / \varepsilon)$ time and space.

\subsection{Dynamic program based on application expenditures}
The first DP produces an optimal solution to the college application problem in $O(Hm + m\log m)$ time. Because we cannot assume that $H \leq m$, this represents a pseudopolynomial-time solution (Garey and Johnson 1979, \S\,4.2). However, it is quite effective for typical instances in which the application costs are small integers. It is also a polynomial-time algorithm for the special case of the college application problem in which each $g_j = 1$, meaning that $H$ is a limit on the \emph{cardinality} $\mathcal{X}$, and $H < \sum g_j = m$ in any nontrivial instance. The algorithm resembles a familiar DP algorithm for the binary knapsack problem (Dantzig 1957).

For $j = 0 \dots m$ and $h = 0 \dots H$, let $\mathcal{X}(j, h)$ denote the optimal portfolio that uses only the schools $\{ 1, \dots, j\}$ and costs no more than $h$, and let $V(j,h) = v\bigl(\mathcal{X}(j, h)\bigr)$. Clearly, if $j=0$ or $h=0$, then $\mathcal{X}(j, h) = \O$ and $V(j, h) = 0$. It is also convenient to let $V(j, h) = -\infty$ for all $h < 0$.

For the remaining indices, $\mathcal{X}(j, h)$ either contains $j$ or not. If it does not contain $j$, then $\mathcal{X}(j, h) = \mathcal{X}(j-1, h)$. On the other hand, if  $\mathcal{X}(j, h)$ contains $j$, then its valuation is $(1 - f_j) v(\mathcal{X}[j, h]\setminus \{j\}) + f_j t_j$. Therefore, $\mathcal{X}(j, h)\setminus \{j\}$ must make optimal use of the remaining budget over the remaining schools; that is, $\mathcal{X}(j, h) = \mathcal{X}(j-1, h - g_j) \cup\{j\}$. From these observations, we obtain the following recursion for $ j = 1\dots m$ and $h = 1\dots H$:
\begin{align}
V(j, h) = \max\bigl\{ V(j-1, h), (1 - f_j) V(j-1, h-g_j) + f_j t_j \bigr\}
\end{align}
with the convention that $ -\infty \cdot 0 = -\infty$. Given the values of $V(j, h)$ at each index, the corresponding optimal portfolios are computed by observing that $\mathcal{X}(j, h)$ contains $j$ if and only if $V(j, h) > V(j-1, h)$. The optimal solution is given by $\mathcal{X}(m, H)$. The algorithm below performs these computations and outputs the optimal portfolio $\mathcal{X}$.

\begin{algorithm}[h] 
%\DontPrintSemicolon
\caption{Application expenditures DP.} \label{ellisDP1}
\KwIn{$f \in \mathbb{Q}^m$, $t \in \mathbb{N}^m$, $g \in \mathbb{N}^m$, $H \in\mathbb{N}$.}
Sort schools by $t_j$ ascending\;
Fill a lookup table with the values of $V(j, h)$\; \label{Vcreatedlookuptable}
$h \gets H$\;
$\mathcal{X} \gets \O$\;
\For{$j = m, m-1, \dots, 1$}{
	\If{$V[j-1, h] < V[j, h]$}{
		$\mathcal{X} \gets \mathcal{X}\cup\{j\}$\; 
		$h \gets h - g_j$\;
	}
}
\Return{$\mathcal{X}$}
\end{algorithm}

\begin{theorem}[Validity of Algorithm \ref{ellisDP1}]
Algorithm \ref{ellisDP1} produces an optimal application portfolio for the college application problem in $O(H m + m \log m)$ time.
\end{theorem}

\begin{proof}
Optimality follows from the foregoing discussion. Sorting $t$ is $O(m \log m)$. The bottleneck step is the creation of the lookup table for $V(j, h)$ in line \ref{Vcreatedlookuptable}. Each entry is generated in unit time, and the size of the table is $O(Hm)$.
\end{proof}

\subsection{Dynamic program based on portfolio valuations}

As with the knapsack problem, the college application problem admits a complementary DP that iterates on the value of the cheapest portfolio instead of on the cost of the most valuable portfolio.
